Local PCA Shows How Population Structure Differs Along the Genome

### 2016-07-07 writeup/local_pca_paper.tex
update

\paragraph{\textit{Drosophila melanogaster}:}
We also used whole-genome sequencing data from Drosophila Population Genomics Project (DPGP) \citep{lack2015drosophila}, 
which together has 380 samples from 16 countries across Africa and Europe.
\plr{Check: are all of our samples described in \citet{lack2015drosophila}? If yes, just say DPGP; if no, add appropriate citation (ask John what this should be). Also, add website?}
Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
regions with residual heterozygosity were marked as missing in the original dataset.
Each chromosome arm we investigated (Chr2L, Chr2R, Chr3L, and Chr3R) has 2--3 million SNPs.
Due to high density of missing data for some parts in the genome, 
we first deleted any samples with more than 8\% missing genotypes, 
and then deleted positions with more than 20\% missing data. 
(We chose these cutoffs as the tails of the relevant empirical distributions.)

@@

After recoding, we divided the genome into contiguous segments (``windows''; see below),
% recoded matrix into contigous matrices that have the same columns but fewer rows than the original matrix, 
and applied Principal Component Analysis (PCA) as described in \citet{mcvean2009genealogical}
separately to the submatrices that corresponded to each window.
Specifically, we did PCA as follows:
denote by $Z$ the $L\times N$ recoded genotype matrix ($L$ is the number of SNPs and $N$ is the sample size), 
and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
so that $\overline{Z_{s}}=\frac{1}{n_s}\sum_j Z_{sj}$, 
where the sum is over the $n_s$ nonmissing genotypes.
We first compute the mean-centered matrix $X$, as $X_{si}=Z_{si}-\overline{Z_{s}}$,
and preserving missingness.
(This mean-centering makes the result not depend on the choice of reference allele,
exactly if there is no missing data, and approximately otherwise.)
Next, we find the covariance matrix of $X$, denoted $C$,
as $C_{ij} = \frac{1}{m_{ij}-1} \sum_s X_{si} X_{sj} - \frac{1}{m_{ij}(m_{ij}-1)} (\sum_s X_{si})(\sum_s X_{sj})$,
where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
\plr{Is that correct? This is not $X^T X/(m-1)$, which is what was here before, 
     but is what cov(), use='pairwise' does: see \url{package/tests/testthat/test_covariance.R}.}
% We compute the covariance matrix using the R function cov() , with use=``pairwise",
% which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
The principal components are the eigenvectors of $C$, 
normalized to have Euclidean length equal to one,
and ordered by magnitude of the eigenvalues.

@@

\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances. Do you think it makes sense here?}
Since eigenvectors are still only defined up to sign,
when comparing between windows we choose the sign to best match each other:
after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\sum_i (\pcone_{i1} - u_i)^{2}$ or
$\sum_i (\pcone_{i1} + u_i)^{2}$ 
is smaller.

@@

We think of local population structure as being summarized by \emph{relative} position of the sampmles
in the space defined by the top principal components.
Since rotations of this space imply identical population structure,
we do not compare population structure of different genomic windows using the PCs directly, 
but instead compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
We do this, rather than using the entire covariance matrix, 
both for computational efficiency 
and because the top PCs summarize important population structure, 
so using only these should reduce the effect of noise.
To remove the effect of variation in mutation rate between windows,
we also rescale each covariance matrix so that the underlying data matrix has trace norm equal to one.
To do this, define the $N \times k$ matrix $V(i)$ so that the $\ell^\text{th}$ column of $V(i)$
is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, $k$-dimensional covariance matrix for the $i^\text{th}$ window is
\begin{align}
    M(i) &= \sum_{\ell=1}^k V_{i\ell} V_{i\ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}

@@

We ran the method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
For each, the two-dimensional MDS visualization resembles a triangle,
as seen in Figure \ref{fig:mds_chr2L}.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme types of population structure 
typified by windows found in the three corners of the triangle,
and that other windows' population structure may be a mixture of those extremes.

\begin{figure}
    \begin{center}
       \includegraphics{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
    \end{center}
    \caption{
        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows,
         and the second and third columns show the position (midpoint) of each window along the genome
         against the first and second MDS coordinates, respectively. 
         Rows show results from different chromosome arms, from top to bottom these are 2L, 2R, 3L, 3R, and X. 
         Colors are consistent for plots in each row. 
         Vertical black lines show the breakpoints of known polymorphic inversions
         (from top to bottom: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
         \label{fig:mds_chr2L}
    }
\end{figure}

To investigate these extremes, for each chromosome arm we visually selected three ``extreme'' windows in the MDS plot,
and take out the 5\% of windows that are closest to it in the MDS coordinates;
highlighted these windows' positions along the genome (colors in Figure \ref{fig:mds_chr2L}),
and produced PCA plots for each collection of windows
(shown in Figure \ref{fig:pca_by_pop}).

\begin{figure}
    \begin{center}
       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L},
        on each chromosome arm of \textit{Drosophila melanogaster}.
        In all plots, each point represents a sample. 
        The first column shows the combined PCA plot for windows 
        whose points are colored green in Figure \ref{fig:mds_chr2L}; 
        the second is for orange windows; 
        and third is for purple windows.
         \label{fig:pca_by_pop}
    }
\end{figure}

The variation in population structure turns out to be almost entirely explained by
several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
obtaining Figure \ref{fig:pca_by_inversion}.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme population structure
shown in green in the first row of Figure \ref{fig:mds_chr2L}
lie directly around the breakpoints of the inversion In(2L)t,
and the PCA plots in the first rows of figure \ref{fig:pca_by_inversion}
shows that population structure here is mostly determined by inversion orientation.
The regions shown in purple lie near the centromere,
and have population structure reflective of two axes of variation,
seen in figure \ref{fig:pca_by_pop},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{pool2015natural} for more details).
\plr{check that reference}
The regions shown in orange mostly lie inside the inversion,
and show population structure that is a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{kirkpatrick2015chromosome}.
\plr{check that reference too}
Similar results are found in other chromosome arms,
\plr{add chromosome X}
except that the situation of Chr3R is a little more complicated,
due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
although we have not highlighted portions of the MDS plot that pick these out.
Two further inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
and do not significantly influence population structure.

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
         In(2L)t, In(2R)NS, In(3L)OK and In(3R)K respectively. 
         (Data from \citet{lack2015drosophila}.
        \label{fig:pca_by_inversion}
    }
\end{figure}

@@

As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
with results shown in Supplementary Figure XXX.
\plr{Add plots for remaining autosomes to supplement.}
On each, variation in population structure was dominated by a small number of windows
having similar population structure to each other that differed dramatically from the rest of the chromosome.
These may be primarily inversions:
for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) 
coincide with the position of a particularly large, polymorphic inversion on 8p23. 
More generally, we see signals coincident with 
three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
shown in Figure \ref{fig:mds_human}.
PCA plots of these windows show a characteristic trimodal shape,
presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
(although we do not have data on orientation status),
a signal that has been proposed as a method to identify inversions \citep{ma2012investigation}.
However, distinguishing this hypothesis from others,
such as regions of low recombination rate,
would require additional data.

We also ran the method on all 22 autosomes together, 
and found that, remarkably, 
the inversion on chromosome 8 is still the most striking outlying signal.
Further investigation with a denser set of SNPs,
allowing a finer genomic resolution,
may yield other patterns.

@@

Across chromosomes, the high values of MDS coordinate 1 coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection \citep{kulikova2001integration,paape2013selection}. 
To verify that this is a possible explanation,
we computed gene density near each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
\plr{Say exactly what was computed.}
These values are shown juxtaposed with
the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
Other genomic features covary with gene density, however,
so we cannot rule out alternative hypotheses.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig5_MDS_and_gene_count_allchr}
    \end{center}
    \caption{
         MDS visualization and gene density for each window in the \textit{Medicago} genome,
         for chromosomes 1--8 (numbered above each pair of figures).
         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
         The black plot below is gene count for each window against the position of each window.
         \label{fig:mds_medicago}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm}
    \end{center}
    \caption{
        First MDS coordinate against gene density for all 8 chromosomes. 
        The first MDS coordinate is significantly correlated with gene count ($r=0.146$, $p=2.2\times 10^{-16}$). 
        (See supplementary for each single chromosome's MDS result against gene count for \textit{Medicago})
        \label{fig:mds_gene_count}
    }
\end{figure}

@@

Our investigations have found a remarkable amount of variation in population structure across the genomes
of three diverse species,
revealing distinct biological processes driving this variation in each species.
More investigation, particularly on more species and datasets will help uncover which patterns are generalizable.

@@

\paragraph{Confounding in GWAS}
So-called cryptic relatedness between samples
has been one of the major sources of confounding in genome-wide association studies (GWAS)
and so methods must account for it by modeling population structure or kinship \citep{gwas_confounding_review,mixed_models}.
Since population structure is not constant along the genome,
this could in principle lead to an inflation of false positives parts of the genome
with stronger population structure than the genome-wide average.
Fortunately, in our human dataset this does not seem likely to have a strong effect:
most variation is due to small, independent regions, possibly primarily inversions,
and so may not have a major effect on GWAS.
In the other species we examined, particularly \textit{Drosophila melanogaster},
treating population structure as a single quantity could be severely misleading.

\paragraph{Parameter choices}
There are several choices in the method that may affect the results.
As with whole-genome PCA,
the choice of samples is important,
as variation not strongly represented in the sample will not be discovered.
The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
but downweighting may be a better option that does not discard data.
\plr{and we have methods to do this}
Next, the choice of window size may be important,
although in our applications results were not sensitive to this,
indicating that the limit of resolution was smaller than the scale on which patterns of kinship varies along the genome.
Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in figure \ref{fig:diagram}),
along with the method used to discover common structure,
will affect results.
We used MDS, applied to either each chromosome separately or to the entire genome;
for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
but genome-wide, their signal is obscured by the numerous other signals of comparable strength.

@@

More generally, there are many possible methods to discover common structure in different parts of the genome.
The methods we chose discovered strong biological signal of different types in three datasets;
but it is possible that other methods for measuring dissimilarity between windows' covariance matrices
or for summarizing the matrix of pairwise distances between windows
would lead to different insights.
Minor points we have not explored include how to decide how many PCs to use in approximating structure of each window
(equations XX),
\plr{(say we have explored)}
how many MDS coordinates to use when describing the distance matrix between windows,
or how to choose interesting regions of the genome when the MDS plot is not triangular.
These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
we encourage the user to experiment.

\paragraph{Chromosomal inversions}
A major driver of variation in population structure in two datasets we examined are inversions.
This may be common,
but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
However, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
or simply regions of reduced recombination.
However, in many applications, inversions are a nuisance.
For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
by regressing out the effect of linked SNPs on each other.
It would be interesting to see if smoehow removing the effects of inversions in the \textit{Drosophila melanogaster} or human datasets
would produce a pattern similar to that seen in \textit{Medicago truncatula}.

@@

>>>>

\paragraph{\textit{Drosophila melanogaster}:}
We also used whole-genome sequencing data from Drosophila Population Genomics Project (DPGP) \citep{lack2015drosophila}, 
which together has 380 samples from 16 countries across Africa and Europe.
Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
regions with residual heterozygosity were marked as missing in the original dataset.
Each chromosome arm we investigated (Chr2L, Chr2R, Chr3L, and Chr3R) has 2--3 million SNPs.
Due to high density of missing data for some parts in the genome, 
we first deleted any samples with more than 8\% missing genotypes, 
and then deleted positions with more than 20\% missing data. 
(We chose these cutoffs as the tails of the relevant empirical distributions.)

@@

After recoding, we divided the genome into contiguous segments (``windows''; see below),
% recoded matrix into contigous matrices that have the same columns but fewer rows than the original matrix, 
and applied Principal Component Analysis (PCA) as described in \citet{mcvean2009genealogical}
separately to the submatrices that corresponded to each window.
Specifically, we did PCA as follows:
denote by $Z$ the $L\times N$ recoded genotype matrix ($L$ is the number of SNPs and $N$ is the sample size), 
and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
so that $\overline{Z_{s}}=\frac{1}{n_s}\sum_j Z_{sj}$, 
where the sum is over the $n_s$ nonmissing genotypes.
We first compute the mean-centered matrix $X$, as $X_{si}=Z_{si}-\overline{Z_{s}}$,
and preserving missingness.
(This mean-centering makes the result not depend on the choice of reference allele,
exactly if there is no missing data, and approximately otherwise.)
Next, we find the covariance matrix of $X$, denoted $C$,
as $C_{ij} = \frac{1}{m_{ij}-1} \sum_s X_{si} X_{sj} - \frac{1}{m_{ij}(m_{ij}-1)} (\sum_s X_{si})(\sum_s X_{sj})$,
where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
% We compute the covariance matrix using the R function cov() , with use=``pairwise",
% which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
The principal components are the eigenvectors of $C$, 
normalized to have Euclidean length equal to one,
and ordered by magnitude of the eigenvalues.

@@

Since eigenvectors are still only defined up to sign,
when comparing between windows we choose the sign to best match each other:
after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\sum_i (\pcone_{i1} - u_i)^{2}$ or
$\sum_i (\pcone_{i1} + u_i)^{2}$ 
is smaller.

@@

We think of local population structure as being summarized by \emph{relative} position of the sampmles
in the space defined by the top principal components.
Since rotations of this space imply identical population structure,
we do not compare population structure of different genomic windows using the PCs directly, 
but instead compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
We do this, rather than using the entire covariance matrix, 
both for computational efficiency 
and because the top PCs summarize important population structure, 
so using only these should reduce the effect of noise.
To remove the effect of variation in mutation rate between windows,
we also rescale each covariance matrix so that the underlying data matrix has trace norm equal to one.
To do this, define the $N \times k$ matrix $V(i)$ so that the $\ell^\text{th}$ column of $V(i)$
is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, $k$-dimensional covariance matrix for the $i^\text{th}$ window is
\begin{align} \label{eqn:est_cov}
    M(i) &= \sum_{\ell=1}^k V_{i\ell} V_{i\ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}

@@

We ran the method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
For each, the two-dimensional MDS visualization resembles a triangle,
as seen in Figure \ref{fig:mds_allchr}.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme types of population structure 
typified by windows found in the three corners of the triangle,
and that other windows' population structure may be a mixture of those extremes.

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
    \end{center}
    \caption{
        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows, 
         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
         rows correspond to chromosome arms.
         Colors are consistent for plots in each row. 
         Vertical lines show the breakpoints of known polymorphic inversions.   
         Solid black lines are for the inversions we used to obtaining Figure \ref{fig:pca_by_inversion},
         while dotted grey lines are for other known inversions.     
         \label{fig:mds_allchr}
    }
\end{figure}

To investigate these extremes, for each chromosome arm we visually selected three ``extreme'' windows in the MDS plot,
and take out the 5\% of windows that are closest to it in the MDS coordinates;
these three groups of windows are shown in color in Figure \ref{fig:mds_allchr} and PCA plots for each collection are shown in Figure \ref{fig:pca_by_pop}.

\begin{figure}
    \begin{center}
       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_allchr},
        on each chromosome arm of \textit{Drosophila melanogaster}.
        In all plots, each point represents a sample. 
        The first column shows the combined PCA plot for windows 
        whose points are colored green in Figure \ref{fig:mds_allchr}; 
        the second is for orange windows; 
        and third is for purple windows.
         \label{fig:pca_by_pop}
    }
\end{figure}

The variation in population structure turns out to be almost entirely explained by
several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
obtaining Figure \ref{fig:pca_by_inversion}.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme population structure
shown in green in the first row of Figure \ref{fig:mds_allchr}
lie directly around the breakpoints of the inversion In(2L)t,
and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
shows that population structure here is mostly determined by inversion orientation.
The regions shown in purple on chromosome 2L lie near the centromere,
and have population structure reflective of two axes of variation,
seen in Figure \ref{fig:pca_by_pop},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{pool2015natural} for more details).
\plr{check that reference}
The regions shown in orange on chromosome 2L mostly lie inside the inversion,
and show population structure that is a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{kirkpatrick2015chromosome}.
\plr{check that reference too}
Similar results are found in other chromosome arms,
except that the situation of Chr3R is a little more complicated,
due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
although we have not highlighted portions of the MDS plot that pick these out.
Two more inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
and do not significantly influence population structure.

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
         (Data from \citet{lack2015drosophila}.
        \label{fig:pca_by_inversion}
    }
\end{figure}

@@

As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
with results shown in Supplementary Figure \ref{fig:mds12_chr1_8_human}, \ref{fig:mds12_chr9_16_human}, \ref{fig:mds12_chr17_22_human}.
On each, variation in population structure was dominated by a small number of windows
having similar population structure to each other that differed dramatically from the rest of the chromosome.
These may be primarily inversions:
for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
coincide with the position of a particularly large, polymorphic inversion on 8p23. 
More generally, we see signals coincident with 
three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
shown in Figure \ref{fig:mds_human}.
PCA plots of these windows show a characteristic trimodal shape (Figure \ref{fig:pca_chr8_outliers}),
presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
(although we do not have data on orientation status),
a signal that has been proposed as a method to identify inversions \citep{ma2012investigation}.
However, distinguishing this hypothesis from others,
such as regions of low recombination rate,
would require additional data.

We also ran the method on all 22 autosomes together, 
and found that, remarkably, 
the inversion on chromosome 8 is still the most striking outlying signal (Figure \ref{fig:mds1_along_allchr_human}). 
Further investigation with a denser set of SNPs,
allowing a finer genomic resolution,
may yield other patterns.

@@

Across chromosomes, the high values of MDS coordinate 1 coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection \citep{kulikova2001integration,paape2013selection}. 
To verify that this is a possible explanation,
we computed gene density for each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
we downloaded the gene model track from Mt4.0 JBrowse at jcvi.org,
extracted the start and end positions of each predicted mRNA transcript,
counted how many of these fell in each window, 
and use the mean number of the starts and ends in a window to represent the gene count for that window.
The endpoints of the windows were defined to be halfway between the last SNP of each window and the first SNP in the next window.
These values are shown juxtaposed with
the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
Other genomic features covary with gene density, however,
so we cannot rule out alternative hypotheses.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig5_MDS_and_gene_count_allchr_update}
    \end{center}
    \caption{
         MDS visualization and gene density for each window in the \textit{Medicago} genome,
         for chromosomes 1--8 (numbered above each pair of figures).
         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
         The black plot below is gene count for each window against the position of each window.
         \label{fig:mds_medicago}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics{MDS_1D_against_gene_count_all_chr_win104_with_lm_update}
    \end{center}
    \caption{
        First MDS coordinate against gene density for all 8 chromosomes. 
        The first MDS coordinate is significantly correlated with gene count ($r=0.149$, $p=2.2\times 10^{-16}$). 
        (See supplementary for each single chromosome's MDS result against gene count for \textit{Medicago})
        \label{fig:mds_gene_count}
    }
\end{figure}

@@

Our investigations have found a remarkable amount of variation in population structure across the genomes
of three diverse species,
revealing distinct biological processes driving this variation in each species.
More investigation, particularly on more species and datasets, will help to uncover which patterns are generalizable.

@@

\paragraph{Confounding in GWAS}
So-called cryptic relatedness between samples
has been one of the major sources of confounding in genome-wide association studies (GWAS)
and so methods must account for it by modeling population structure or kinship \citep{gwas_confounding_review,mixed_models}.
Since population structure is not constant along the genome,
this could in principle lead to an inflation of false positives in parts of the genome
with stronger population structure than the genome-wide average.
Fortunately, in our human dataset this does not seem likely to have a strong effect:
most variation is due to small, independent regions, possibly primarily inversions,
and so may not have a major effect on GWAS.
In the other species we examined, particularly \textit{Drosophila melanogaster},
treating population structure as a single quantity could be severely misleading.

\paragraph{Parameter choices}
There are several choices in the method that may affect the results.
As with whole-genome PCA,
the choice of samples is important,
as variation not strongly represented in the sample will not be discovered.
The effects of strongly imbalanced sampling schemes are often corrected by dropping samples in overrepresented groups;
but downweighting may be a better option that does not discard data.
\plr{and we have methods to do this}
Next, the choice of window size may be important,
although in our applications results were not sensitive to this,
indicating that we can see variation on a sufficiently fine scale.
Finally, which collections of genomic regions are compared to each other (steps 3 and 4 in Figure \ref{fig:diagram}),
along with the method used to discover common structure,
will affect results.
We used MDS, applied to either each chromosome separately or to the entire genome;
for instance, human inversions are clearly visible as outliers when compared to the rest of their chromosome,
but genome-wide, their signal is obscured by the numerous other signals of comparable strength.

@@

More generally, there are many possible methods to discover common structure in different parts of the genome.
The methods we chose discovered strong biological signal of different types in three datasets;
but it is possible that other methods for measuring dissimilarity between windows' covariance matrices
or for summarizing the matrix of pairwise distances between windows
would lead to different insights.
Minor points we have not explored include how to decide how many PCs to use in approximating structure of each window
(equations \ref{eqn:est_cov}),
\plr{(say we have explored)}
how many MDS coordinates to use when describing the distance matrix between windows,
or how to choose interesting regions of the genome when the MDS plot is not triangular.
These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
we encourage the user to experiment.

\paragraph{Chromosomal inversions}
A major driver of variation in population structure in two datasets we examined are inversions.
This may be common,
but the example of \textit{Medicago truncatula} shows that polymorphic inversions are not ubiquitous.
PCA has been proposed as a method for discovering inversions \citep{ma2012investigation};
However, the signal left by inversions likely cannot be distinguished from long haplotypes under balancing selection 
or simply regions of reduced recombination.
However, in many applications, inversions are a nuisance.
For instance, SMARTPCA \citep{patterson2006population} reduces their effect on PCA plots
by regressing out the effect of linked SNPs on each other.
It would be interesting to see if somehow removing the effects of inversions in the \textit{Drosophila melanogaster} or human datasets
would produce a pattern similar to that seen in \textit{Medicago truncatula}.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr1_8_all_valid_and_pre}
    \end{center}
    \caption{
         MDS plots for human chromosome 1-8.
         The first column shows the MDS visualization of relationships between windows, 
         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
         rows correspond to chromosomes.        
         Colorful vertical lines show the breakpoints of known valid inversions, 
         while grey vertical lines show the breakpoints of predicted inversions.
        \label{fig:mds12_chr1_8_human}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr9_16_all_valid_and_pre}
    \end{center}
    \caption{
        MDS plots for human chromosome 9-16.
        \label{fig:mds12_chr9_16_human}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{FigS_human_MDS12_chr17_22_all_valid_and_pre}
    \end{center}
    \caption{
        MDS plots for human chromosome 17-22.
        \label{fig:mds12_chr17_22_human}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{FigS_PCA_chr8_outliers_combine}
    \end{center}
    \caption{
        PCA plots comparison for outlier windows and non-outlier windows
        \label{fig:pca_chr8_outliers}
    }
\end{figure}
